---
title: metaboWorkflows
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include=FALSE,echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, 
                      comment = "#>",
                      fig.align = 'center',
                      fig.path = "man/figures/README-",
                      message = FALSE)
library(metaboWorkflows)
```

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![R-CMD-check](https://github.com/jasenfinch/metaboWorkflows/workflows/R-CMD-check/badge.svg)](https://github.com/jasenfinch/metaboWorkflows/actions)
[![Codecov test coverage](https://codecov.io/gh/jasenfinch/metaboWorkflows/branch/master/graph/badge.svg)](https://codecov.io/gh/jasenfinch/metaboWorkflows?branch=master)
[![license](https://img.shields.io/badge/license-GNU%20GPL%20v3.0-blue.svg)](https://github.com/jasenfinch/metaboWorkflows/blob/master/DESCRIPTION) 
<!-- badges: end -->

> Workflow Project Templates for Metabolomics Analyses

## Overview

This package provides reproducible workflow project templates for metabolomics analyses using the [hrm](https://jasenfinch.github.io/hrm/) R package ecosystem.

These project templates utilise a number of tools to promote efficient and reproducible analysis, agnostic of the actual analysis R code.
These tools include:

* [targets](https://docs.ropensci.org/targets/) - an R focused pipeline toolkit for efficiently maintaining reproducible analysis workflows.
* [renv](https://rstudio.github.io/renv/) - an R package for project-local R package dependency management for maintaining reproducible R package environments.
* [git](https://git-scm.com/) - a widely used, open-source distributed version control system.
* [docker](https://www.docker.com/) - enables the containerization of operating system (OS) level environments. This can be used to define reproducible OS environments in which a workflow analysis can be performed.

Project templates are available for the following metabolomic techniques:

```{r available_workflows,eval=TRUE,echo=FALSE,results='asis'}
availableWorkflows() %>% 
  {paste0('* ',.,collapse = '\n')} %>% 
  cat()
```

## Installation

The `metaboWorkflows` package can be installed from GitHub using the following:

```r
devtools::install_github('jasenfinch/metaboWorkflows',build_vignettes = TRUE)
```

## Learn more

The package documentation can be browsed online at <https://jasenfinch.github.io/metaboWorkflows/>. 

If this is your first time using `metaboWorkflows` see the [Introduction](https://jasenfinch.github.io/metaboWorkflows/articles/metaboWorkflows.html) vignette for information on how to get started.

If you believe you've found a bug in `metaboWorkflows`, please file a bug (and, if
possible, a [reproducible example](https://reprex.tidyverse.org)) at
<https://github.com/jasenfinch/metaboWorkflows/issues>.
