---
title: "Introduction"
subtitle: "`r paste0('metaboWorkflows v',packageVersion('metaboWorkflows'))`"
author: "Jasen Finch"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  prettydoc::html_pretty:
    toc: true
    highlight: github
    theme: tactile
vignette: >
  %\VignetteIndexEntry{Introduction to metaboWorkflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r example-data-check, include=FALSE}
available_datasets <- metaboData::availableDataSets()

if (available_datasets$status[available_datasets$technique == 'FIE-HRMS' & 
                              available_datasets$`data set` == 'BdistachyonEcotypes'] != 'available'){
  metaboData::downloadDataSet('FIE-HRMS','BdistachyonEcotypes',ask = FALSE)
}
```

## Introduction

Metabolomics encompasses a wide range of techniques that includes mass spectrometry based fingerprinting and profiling.
Analysis of the this data requires number of steps that includes spectral processing, data pre-treatment, quality control, data mining and visualisation in order to extract relevant biological information.
The goal of `metaboWorkflows` is to provide project directory templates for a range of mass spectrometry based techniques.
This leverages a number of tools to promote efficient and reproducible analysis, agnostic of the actual analysis R code.
These tools include:

* [targets](https://docs.ropensci.org/targets/) - an R focused pipeline toolkit for efficiently maintaining reproducible analysis workflows.
* [renv](https://rstudio.github.io/renv/) - an R package for project-local R package dependency management for maintaining reproducible R package environments.
* [git](https://git-scm.com/) - a widely used, open source distributed version control system.
* [docker](https://www.docker.com/) - enables the containerization of operating system (OS) level environments. This can be used to define reproducible OS environments in which a workflow analysis can be performed.

While the use of most of these tools is optional but highly encouraged, it is recommended for the user to at least be familiar with the basic use of `targets` based workflows outlined [here](https://books.ropensci.org/targets/walkthrough.html).

There are three steps in using `metaboWorkflows` to generate a workflow project template directory. 
These are as follows:

* **input** - define the input data source of the workflow
* **define** - define the workflow steps based on the metabolomic technique
* **generate** - generate the workflow project directory

This introduction will outline each of these steps and provided an overview of an example workflow project directory along with how to execute one of these workflows.

To begin, first load the package:

```{r setup}
library(metaboWorkflows)
```

## Workflow input

Prior to defining a workflow, the user needs to know what sort of input the type the workflow will use.
`metaboWorkflows` currently supports two types: use of remote data obtained through the use of a [`grover`](https://jasenfinch.github.io/grover/) API or through providing a vector .mzML file paths and a [`tibble`](https://tibble.tidyverse.org/) of sample information.

### `grover` API input

The [`grover`](https://jasenfinch.github.io/grover/) R provides a framework for hosting web APIs for remote access and conversion of raw metabolomics data to mzML format.

This type of input can be declared by providing the host information of the [`grover`](https://jasenfinch.github.io/grover/) API to the `inputGrover` function.
Below shows an example for a ficticious [`grover`](https://jasenfinch.github.io/grover/) API host.

```{r grover-input}
workflow_input <- inputGrover(instrument = 'An_instrument',
                              directory = 'Experiment_directory',
                              host = 'a.grover.host',
                              port = 80,
                              auth = '1234')
```

```{r print-grover-input}
print(workflow_input)
```

### File path input

If the raw mass spectrometry .mzML format data files are already availble for a sample set, their file paths and a [`tibble`](https://tibble.tidyverse.org/) of the sample information can be provided for workflow input.

Below shows an example file path input for the FIE-HRMS fingerprinting data set of *Brachypodium distachyonn* ecotype comparisons available in the [`metaboData`](https://aberhrml.github.io/metaboData/) package.

```{r file-path-input}
file_paths <- metaboData::filePaths('FIE-HRMS','BdistachyonEcotypes')
sample_information <- metaboData::runinfo('FIE-HRMS','BdistachyonEcotypes')

workflow_input <- inputFilePath(file_paths,sample_information)
```

```{r print-file-path-input}
print(workflow_input)
```

This example input will be used throughout the rest of this introduction.

## Define a workflow

Workflow definition is simple and requires the input definition outlined in the previous section, the name of the metabolomic technique and the project name.

To return the metabolomic workflow templates that are currently available, use:

```{r available-workflows}
availableWorkflows()
```

This example uses the `FIE-HRMS fingerprinting` template and the project will be named `Example project`. 

```{r workflow-definition}
workflow_definition <- defineWorkflow(
  input = workflow_input,
  workflow = 'FIE-HRMS fingerprinting',
  project_name = 'Example project'
)
```

Printing the resulting workflow definition will provide information about the workflow.

```{r print-workflow}
print(workflow_definition)
```

As shown above, this workflow definition contains `r targets(workflow_definition) %>% unlist(recursive = TRUE) %>% length()` targets.
These targets are the individual steps in the pipeline.
The full workflow graph, showing the relationships between these targets, can be plotted as shown below:

```{r glimpse-workflow}
glimpse(workflow_definition)
```

The individual workflow targets are stored as lists, nested into a number of modules.
The modules available within a given workflow can be viewed:

```{r workfow-modules}
modules(workflow_definition)
```

The individual analysis targets can also be viewed.
The following shows the nested list of targets found within the input module.

```{r input-targets}
targets(workflow_definition)$input
```

These workflow targets are fully customisable which is outlined in the [Workflow customisation](customisation.html) vignette.

## Generate a workflow project directory

The final step is the generation of the project directory for the defined workflow.

```{r generate-workflow,eval=FALSE}
generateWorkflow(workflow_definition)
```

## The project directory

```
Example_project/
├── Dockerfile
├── Example_project.Rproj
├── R
│   ├── functions
│   └── utils.R
├── README.md
├── _targets.R
├── data
│   ├── file_paths.txt
│   └── runinfo.csv
├── exports
├── misc
├── renv
│   ├── activate.R
│   ├── library
│   ├── local
│   ├── settings.dcf
│   └── staging
├── renv.lock
└── run.R
```

## Executing an analysis workflow

